---
title: "Getting started with AttrMort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with AttrMort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

`AttrMort` estimates the deaths attributable to ambient air pollution. It is
**pollutant-agnostic** -- the built-in concentration-response (C-R) models
cover PM<sub>2.5</sub>, O<sub>3</sub> and NO<sub>2</sub>, and anything else can
be supplied as a lookup table or as a JSON configuration -- and
**input-agnostic**: data frames, CSV/Excel files, gridded GeoTIFF/netCDF
rasters and spatial vector layers all go through the same pipeline.

```{r library}
library(AttrMort)
```

This vignette walks through

1. the data contract,
2. one attribution run, argument by argument,
3. domain-level results and the uncertainty range,
4. a two-group decomposition of a change,
5. raster and vector inputs, and one grid reused across scenarios,
6. custom concentration-response functions,
7. the exported functions at a glance.

Every example uses the synthetic data shipped in `inst/extdata/`, so the whole
vignette runs offline. The 0.25 degree grid has 6000 cells in three fictional
countries (`Aland`, `Borduria`, `Cyrenia`); the wide tables carry two scenario
columns, `base2015` and `scenario2030`.

## The data contract

| Column | Meaning |
|---|---|
| `conc` | concentration, keyed as a character string at `dgt_conc` decimals |
| `pop` | population count of the cell |
| `age` | age stratum as character five-year groups (`"25"`, `"30"`, ...) |
| `prop` | share of the domain population in that age group |
| `endpoint` | cause label, matched against the C-R table |
| `mortrate` | baseline mortality, deaths per 100,000 |
| `location` | domain label used to join the domain-level tables |

Two contracts matter more than the rest:

* **Keys are strings at a fixed precision.** Coordinate keys are rendered at
  `dgt_coord` decimals (default 2) and concentration keys at `dgt_conc`
  decimals (default 1); `matchable()` is the renderer used on both sides. The
  same rounding has to be applied to the exposure and to the lookup table, or
  the join finds nothing.
* **Mortality is published per 100,000.** The calculation divides by 1e5 once,
  so what comes out is a number of deaths.

Column names are matched heuristically when they are not already canonical:
names are compared after lowercasing and dropping separators (`mort_rate`,
`Mort Rate` and `mortrate` are the same column), and a recognised variant is
renamed to the canonical name (`age_group` to `age`, `mortality_rate` to
`mortrate`, `Endpoint` to `endpoint`). Coordinate columns are renamed to
`x`/`y`, and `lon`/`lat`, `long`/`lat`, `longitude`/`latitude` are all the
same pair -- so two tables that spell their keys differently still join.

## One attribution run, argument by argument

```{r quick-start}
extdata <- system.file("extdata", package = "AttrMort")
sheet <- function(f) readxl::read_excel(file.path(extdata, f))

deaths <- mortality(
  crf       = "GEMM",
  calc_fild = sheet("grid_info.xlsx"),
  conc_real = sheet("grid_exposure.xlsx"),
  pop_total = sheet("grid_pop.xlsx"),
  age_struc = sheet("national_age_structure.xlsx"),
  mort_rate = sheet("national_mortality.xlsx"),
  scenario  = "base2015",
  mort_lvl  = "location"
)

dim(deaths)
deaths[1:3, c("x", "y", "location", "ncd+lri_25")]
```

Each argument does one thing:

* `crf` selects the C-R model. `cr_models()` lists the accepted names and
  aliases; a data.frame is accepted instead when the curve is not one of the
  built-ins (see the last section).
* `ci` would pick the lookup branch (`"MEAN"`, `"UP"`, `"LOW"`; `"UPPER"` and
  `"LOWER"` are aliases). It is ignored when `crf` is a data.frame.
* `calc_fild` is the attribution field: the spatial skeleton -- coordinate
  and/or domain columns -- every other input is joined onto.
* `conc_real` is the exposure; for a raster it also defines the analysis grid.
* `conc_cf` is the counterfactual exposure. `NULL` (the default) reuses
  `conc_real`, which means "no additional burden above the counterfactual".
* `pop_total` is the population; `age_struc` supplies `prop`, the share of each
  age group in each domain (the proportions are renormalised within a domain,
  so counts and percentages are both accepted).
* `mort_rate` is cause-specific baseline mortality in deaths per 100,000.
* `mort_lvl` names the domain used to calibrate the population-weighted
  relative risk (PWRR), here the `location` column. `mort_lvl = NULL` skips
  calibration and computes at the finest available level; that is a different
  quantity, not a numerically identical shortcut.
* `scenario = "base2015"` narrows each wide input to one column. It is a
  **per-input** selector, not a contract across inputs: an input without that
  column is used as it is when it has its canonical value column (`conc`,
  `pop`, `prop`, `mortrate`) or a single numeric value column. The package
  never judges whether the inputs belong to the same scenario or year. Several
  candidates and no canonical column stop the call with the list, rather than
  guessing.
* `dgt_coord` and `dgt_conc` change how the keys are rendered; they have to
  match the precision of the lookup table.
* `target_res` pins the grid resolution of a raster run. `NULL` (the default)
  detects the finest input resolution: grids of up to 5e6 cells are confirmed
  silently, larger ones interactively, and a non-interactive session above 5e7
  cells falls back to 0.1 degree with a message.
* `validate` is `"warn"` (default), `"stop"` or `"off"`. `"off"` is complete
  silence -- preparation messages, mapping notes, validation warnings, the grain
  line, the branch notice and the uncertainty report are all suppressed, and no
  number changes.
* `chunk_ages` sets how many age strata are computed per pass; `NULL` sizes the
  pass from the number of cells, ages and endpoints so that one pass stays
  around 2 GB. Splitting by age is numerically exact, so a chunked run is
  identical to an unchunked one.
* `uncertain` and `conc_uncert` ask for the uncertainty range, `aggregate` and
  `aggregate_by` for domain aggregation; both are the subject of the next two
  sections.

The result is one row per grid cell and one `{endpoint}_{age}` column per
stratum. Endpoint names may contain `+`, `.` or `_` (`ncd+lri`, `copd_lri_v2`),
so a result column is parsed from the right at the last underscore: the
trailing token is the age, everything before it the endpoint.

The run also says what it did. One `Analysis grain:` line reports where the
analysis lives and which calibration branch was taken:

```text
Analysis grain: <n> cell(s) in <m> domain(s) on a <res> deg grid;
PWRR calibrated per domain.
```

The other branch wordings are `grid-level PAF without calibration` (for
`mort_lvl = NULL`) and a one-unit fallback when `mort_rate` carries no domain
column at all -- which also produces a warning, because the whole field is
then calibrated as a single unit. Note that the grid-level and domain-level
grains are not numerically identical: a grid run takes the population-weighted
mean of a domain's cell-level relative risks, a domain-only run the relative
risk at the domain's population-weighted mean concentration.

## Domain-level results and the uncertainty range

`aggregate = TRUE` sums the grid-level burdens per domain, so one row per
domain comes out, with `conc_pwe` -- the population-weighted concentration of
that domain -- beside the deaths. `aggregate_by` keeps a breakdown:

```{r aggregate}
by_location <- mortality(
  crf          = "GEMM",
  calc_fild    = sheet("grid_info.xlsx"),
  conc_real    = sheet("grid_exposure.xlsx"),
  pop_total    = sheet("grid_pop.xlsx"),
  age_struc    = sheet("national_age_structure.xlsx"),
  mort_rate    = sheet("national_mortality.xlsx"),
  scenario     = "base2015",
  mort_lvl     = "location",
  aggregate    = TRUE,
  aggregate_by = "total",
  uncertain    = TRUE,
  conc_uncert  = 12
)

by_location[, c("location", "total", "CI_LOW", "CI_UP", "conc_pwe")]
```

* `aggregate = TRUE` groups by `mort_lvl`. A character vector instead groups
  by exactly those result columns, e.g. `aggregate = "location"`, and with
  `mort_lvl = NULL` the whole field becomes a single row.
* `aggregate_by` is `"total"` (one `total` column), `"endpoint"`
  (`{endpoint}_all` columns), `"age"` (`all_{age}` columns) or `"all"`.
* `uncertain = TRUE` reports `CI_LOW`/`CI_UP` next to the central estimate.
* `conc_uncert = 12` adds an exposure uncertainty of 12 percent.

The interval is a **range, not a sampling interval**. The published low/high
C-R tables are applied to every grid cell and the per-cell burdens are summed.
A C-R function is a single curve shared by all cells, so its uncertainty is
common-mode and does not average out as the domain grows; with one shared
parameter the quantiles of the total are exactly the total evaluated at the
parameter quantiles. `conc_uncert` runs a second chain with every
concentration scaled by `1 +/- percent / 100` and reports the union of the two
ranges (summed the same way, not an independent-error sum). A cell that at
least one chain cannot compute gets `NA` and a warning: its interval is
undefined there, not copied from a neighbour.

The same aggregation is available after the fact, which is convenient when one
grid-level result is reused:

```{r aggregate-post-hoc}
# sum the grid-level result within each location
aggregate_mortality(deaths, at = "location", by = "total")

# keep the endpoint breakdown
aggregate_mortality(deaths, at = "location", by = "endpoint")
```

`aggregate_ci()` does the analogous job for a result whose columns carry a CI
suffix (`_MEAN`, `_UP`, `_LOW`), keeping the three branches side by side:

```{r aggregate-ci}
run_branch <- function(ci) {
  mortality(
    crf = "GEMM", ci = ci,
    calc_fild = sheet("grid_info.xlsx"),
    conc_real = sheet("grid_exposure.xlsx"),
    pop_total = sheet("grid_pop.xlsx"),
    age_struc = sheet("national_age_structure.xlsx"),
    mort_rate = sheet("national_mortality.xlsx"),
    scenario = "base2015", mort_lvl = "location", validate = "off"
  )
}

append_ci <- function(x, suffix) {
  values <- setdiff(names(x), c("x", "y", "location"))
  names(x)[names(x) %in% values] <- paste0(values, suffix)
  x
}

grid_ci <- append_ci(run_branch("MEAN"), "_MEAN")
grid_ci <- dplyr::left_join(
  grid_ci, append_ci(run_branch("UPPER"), "_UP"),
  by = c("x", "y", "location")
)
grid_ci <- dplyr::left_join(
  grid_ci, append_ci(run_branch("LOWER"), "_LOW"),
  by = c("x", "y", "location")
)

aggregate_ci(grid_ci, at = "location", by = "total")
```

Both `aggregate_mortality()` and `aggregate_ci()` accept a named list of
results (one per scenario) and add a `scenario` column to the output.
`write_mortality_xlsx()` writes a data.frame or such a list to a path you
choose; it needs the suggested `writexl` package:

```{r write-xlsx, eval = requireNamespace("writexl", quietly = TRUE)}
path <- write_mortality_xlsx(
  by_location,
  path = tempdir(),
  suffix = "location_total_2015"
)
basename(path)
```

## Comparing two groups: `decompose()`

`decompose()` splits the change in attributable deaths between two complete
sets of inputs -- `from` and `to` -- among four drivers:

| Driver | What it moves |
|---|---|
| `PG` | population growth (`pop_total`) |
| `PA` | population ageing (`age_struc`) |
| `EXP` | the counterfactual exposure (`conc_cf`; the group's own `conc_real` when it has none) |
| `ORF` | the real exposure (`conc_real`) together with the baseline mortality rate |

Each group is a list with one table per role. There is no scenario argument:
the comparison is between two datasets, so each role carries one value column
and the scenario is selected before the call.

```{r decompose}
grid_info <- sheet("grid_info.xlsx")
age_struc <- sheet("national_age_structure.xlsx")
mort_rate <- sheet("national_mortality.xlsx")
exposure  <- sheet("grid_exposure.xlsx")
pop       <- sheet("grid_pop.xlsx")

as_group <- function(scenario, label) {
  list(
    label     = label,
    conc_real = data.frame(exposure[c("x", "y")],
                           conc = exposure[[scenario]]),
    pop_total = data.frame(pop[c("x", "y")],
                           pop = pop[[scenario]]),
    age_struc = data.frame(age_struc[c("location", "age")],
                           prop = age_struc[[scenario]]),
    mort_rate = data.frame(mort_rate[c("location", "age", "endpoint")],
                           mortrate = mort_rate[[scenario]])
  )
}

res <- decompose(
  crf       = "GEMM",
  calc_fild = grid_info,
  mort_lvl  = "location",
  from      = as_group("base2015", "base2015"),
  to        = as_group("scenario2030", "scenario2030")
)

length(res)
names(res)[1:3]
```

A group's optional `conc_cf` is its counterfactual exposure and defaults to
its own `conc_real`; `label` is cosmetic and only names the group in the
summary the call prints.

The return value is a named list of **24 data frames**, one per ordering of
the four drivers. Each holds the key columns, `Cause_Age`, `Start`, one column
per driver **in the order that ordering moves them**, and `End`, with one row
per grid cell and age stratum:

```{r decompose-one}
one <- res[["PG-PA-EXP-ORF"]]
dim(one)
names(one)
head(one, 3)

# per row: Start + the four drivers == End
drivers <- c("PG", "PA", "EXP", "ORF")
max(abs(rowSums(one[drivers]) + one$Start - one$End))
```

A step reads each role from `to` when its driver has already moved and from
`from` otherwise, so a driver column isolates the driver it is named after,
`Start` is the `from` group and `End` the `to` group. The 24 orderings share
**16 states** (one per subset of drivers that has moved), the inputs are
prepared once per group and the C-R table is built once, so a full
decomposition costs sixteen calculations rather than 24 x 5. `Start` and `End`
are identical in every table.

The orderings disagree by construction -- moving the same drivers in a
different order attributes part of the same change to a different driver.
That is why all of them are returned, and why the order-independent reading is
their **mean**:

```{r decompose-mean}
mean_driver <- Reduce(`+`, lapply(res, function(x) x[drivers])) / length(res)

head(mean_driver, 3)

# the mean contribution of each driver over the 24 orderings
round(colSums(mean_driver))
```

A cell and stratum that some state cannot compute (no risk in the C-R table,
no domain label, or the group does not cover it) comes back as `NA` with a
warning, so the affected contribution is visibly missing rather than borrowed
from a neighbouring cell. `validate = "stop"` turns that warning into an
error; `validate = "off"` silences the whole run.

## Raster and vector inputs

A raster exposure already carries the analysis grid, so `calc_fild` can be
left out: `conc_real` defines the target grid, `pop_total` is aggregated onto
it with the total preserved, and boundaries are rasterized onto it. This chunk
turns the example grid into two GeoTIFFs in a temporary directory and runs the
raster-only path:

```{r rasters}
template <- terra::rast(xmin = 5, xmax = 30, ymin = 35, ymax = 50,
                        resolution = 0.25)
centres <- terra::xyFromCell(template, seq_len(terra::ncell(template)))
keys <- paste(round(centres[, 1], 2), round(centres[, 2], 2))

to_raster <- function(df, value) {
  r <- template
  terra::values(r) <- df[[value]][
    match(keys, paste(round(df$x, 2), round(df$y, 2)))
  ]
  r
}

conc_tif <- tempfile(fileext = ".tif")
pop_tif  <- tempfile(fileext = ".tif")
terra::writeRaster(to_raster(exposure, "base2015"), conc_tif)
terra::writeRaster(to_raster(pop, "base2015"), pop_tif)

boundary <- sf::st_sf(
  admin = c("Aland", "Borduria", "Cyrenia"),
  geometry = sf::st_sfc(
    sf::st_polygon(list(rbind(c(5, 35), c(13, 35), c(13, 50),
                              c(5, 50), c(5, 35)))),
    sf::st_polygon(list(rbind(c(13, 35), c(21, 35), c(21, 50),
                              c(13, 50), c(13, 35)))),
    sf::st_polygon(list(rbind(c(21, 35), c(30, 35), c(30, 50),
                              c(21, 50), c(21, 35)))),
    crs = 4326
  )
)

deaths_raster <- mortality(
  crf       = "GEMM",
  conc_real = conc_tif,
  pop_total = pop_tif,
  age_struc = sheet("national_age_structure.xlsx"),
  mort_rate = sheet("national_mortality.xlsx"),
  scenario  = "base2015",
  admin     = boundary,
  mort_lvl  = "location"
)

dim(deaths_raster)
```

Things worth knowing about this path:

* **Resolution.** `target_res` takes one number for a square grid or
  `c(res_x, res_y)` for a non-square one. Rasters are never resampled onto the
  average of the two axes; the resolution is kept per axis. A raster input
  finer than `conc_real` lowers the target resolution, two rasters at
  different resolutions are reconciled by aggregating population with a sum
  (the total is checked, not approximated) before resampling, and a grid that
  would exceed 1e9 cells is refused rather than attempted.
* **Multi-band rasters** (one band per scenario) must share one validity mask.
  A cell missing from some bands is reported with the per-band counts and
  removed from the analysis grid, rather than silently changing the grid
  between scenarios.
* **Boundary sources.** `admin` is an `sf` object or a vector path
  (`.shp`, `.gpkg`, `.geojson`); the map's label column is named by
  `admin_col` (`"admin"` by default, falling back to the map's first character
  column with a warning) and is renamed to the resolved `mort_lvl`. With
  `admin = NULL`, a raster run whose `mort_lvl` is a column of `mort_rate` but
  not of `calc_fild` falls back to national boundaries from
  `rnaturalearth::ne_countries(scale = 110)` and reports how many domain
  names matched -- that fallback needs the suggested `rnaturalearth` package
  and `mort_lvl` values that match its country names, which the fictional
  example data do not, so the run below is not evaluated:

```{r national-default, eval = FALSE}
# needs rnaturalearth and country names matching `mort_lvl`
mortality(
  crf       = "GEMM",
  conc_real = "pm25_base2015.tif",
  pop_total = "population.tif",
  age_struc = "age_structure.csv",
  mort_rate = "mortality.csv",
  scenario  = "base2015",
  mort_lvl  = "location"
)
```

A **table** with coordinate columns can also be used without `calc_fild`: the
skeleton is then the exposure's own cells, and the run reports how many
coordinate pairs it used. The skeleton holds coordinates only, so the
domain-keyed inputs need a domain column -- supplied here through `admin`:

```{r no-calc-fild}
deaths_table <- mortality(
  crf       = "GEMM",
  conc_real = exposure,
  pop_total = pop,
  age_struc = age_struc,
  mort_rate = mort_rate,
  scenario  = "base2015",
  admin     = boundary,
  mort_lvl  = "location"
)

dim(deaths_table)
```

Finally, a **vector map passed as `calc_fild`** is a boundary source too: the
grid still comes from the exposure and the map labels the cells. The two
boundary sources cannot be combined -- passing the map both ways is an error.

```{r map-calc-fild}
deaths_map <- mortality(
  crf       = "GEMM",
  calc_fild = boundary,
  conc_real = exposure,
  pop_total = pop,
  age_struc = age_struc,
  mort_rate = mort_rate,
  scenario  = "base2015",
  mort_lvl  = "location"
)

dim(deaths_map)
```

## One grid, reused

Every scenario of an analysis has to run on the same grid, so the grid is
worth an object of its own. `build_grid_info()` produces it, and it runs the
same alignment, ingestion and boundary rasterization steps a `mortality()`
call runs, so the table describes exactly the grid those calls use:

```{r grid-info}
gi <- build_grid_info(
  conc_real = conc_tif,
  pop_total = pop_tif,
  admin     = boundary,
  path      = file.path(tempdir(), "grid_info.rds")
)

dim(gi)
attr(gi, "res")           # degrees
attr(gi, "ext")           # xmin, xmax, ymin, ymax
attr(gi, "n_cells")
```

The table carries the coordinate keys and the domain labels, never value
columns (`conc`, `pop`, scenario columns), plus the georeferencing as
attributes (`res`, `ext`, `crs`, `n_cells`, `dgt_coord`). Hand it back as
`calc_fild =` next to rasters and `mortality()` checks the keys against the
raster grid: sharing zero coordinate keys is an error, a partial overlap is a
warning with the number of unmatched keys. That is what makes a grid written
for another resolution visible instead of silently narrowing the analysis;
`validate = "off"` disables the check as well.

## Custom concentration-response functions

Three routes lead to a C-R function the package does not ship.

**A data.frame** with the long contract (`conc`, `endpoint`, `age`, `RR`)
is the quickest one:

```{r crf-df}
crf_custom <- data.frame(
  conc     = matchable(seq(0, 60, 1), 1),
  endpoint = "ncd+lri",
  age      = "25",
  RR       = 1 + 0.01 * seq(0, 60, 1)
)

deaths_custom <- mortality(
  crf       = crf_custom,
  calc_fild = grid_info,
  conc_real = exposure,
  pop_total = pop,
  age_struc = age_struc,
  mort_rate = mort_rate,
  scenario  = "base2015",
  mort_lvl  = "location"
)

# only cells whose concentration is inside the lookup range survive
nrow(deaths_custom)
```

Two traps here. The lookup must cover the exposure range: values outside it
are dropped by the join and reported ("`n` value(s) ... fall outside the CRF
lookup range ..."), and when nothing survives the call stops. And the
concentration keys must be rendered the same way on both sides -- the
`dgt_conc` argument of `mortality()` and the `dgt` argument of `rr_std()`
exist to keep them in step.

**A JSON configuration.** `cr_config()` reads the metadata -- which lookup a
model uses, which column carries the concentration axis, which endpoints
apply to which ages -- and `cr_models()` lists the accepted names. The shipped
configuration covers the built-in models:

```{r cr-config}
cr_models()

cfg <- cr_config()
names(cfg$models)
cfg$models$NO2$endpoints[[1]]
```

A config is a JSON object with `schema_version` and `models`; each model has a
`label`, `aliases`, a `lookup`, a `conc_col` and `endpoints`. The `lookup`
kind is `"rda"` (with `table`), `"xlsx"` (with `path` and optionally
`sheets`), or `"csv"` (with `path`, a directory of `MEAN.csv`/`LOW.csv`/
`UP.csv`). Every `endpoint` has a `name` and `ages`, either a range
(`{from, to, by}`) or an explicit numeric vector; a range may ask for at most
100 ages, and more is an error. Relative lookup paths resolve against the
directory of the config file, never the working directory.

The end-to-end route for a new pollutant or endpoint set is: coefficients
into `build_cr_table()`, the workbook into a config, the config into
`mortality()`.

```{r crf-json}
coefs <- data.frame(
  cause = "IHD", age = "25", alpha = 5, beta = 0.01,
  gamma = 0.5, tmrel = 3
)
user_cr <- build_cr_table(
  coefs, model = "IER", conc = seq(0, 300, 0.1),
  path = file.path(tempdir(), "my_cr.xlsx")
)
names(user_cr)

cfg <- list(
  schema_version = 1,
  models = list(
    MYMODEL = list(
      label = "PM2.5_my_model",
      aliases = "MYCURVE",
      lookup = list(kind = "xlsx", path = "my_cr.xlsx",
                    sheets = list(MEAN = "MEAN", LOW = "LOW", UP = "UP")),
      conc_col = "conc",
      endpoints = list(list(name = "ihd",
                            ages = list(from = 25, to = 95, by = 5)))
    )
  )
)
cfg_path <- file.path(tempdir(), "cr_models_my.json")
jsonlite::write_json(cfg, cfg_path, auto_unbox = TRUE, pretty = TRUE)

user_lookup <- rr_std("MYCURVE", "MEAN", config = cfg_path)
head(user_lookup, 3)

deaths_my <- mortality(
  crf       = "MYCURVE",
  cr_config = cfg_path,
  calc_fild = grid_info,
  conc_real = exposure,
  pop_total = pop,
  age_struc = age_struc,
  mort_rate = mort_rate,
  scenario  = "base2015",
  mort_lvl  = "location"
)

names(deaths_my)[2:4]
```

**Coefficients.** `build_cr_table()` implements two published forms, chosen
with `model`:

* `"IER"` (or an IER variant such as `"IER2017"`): one curve per parameter
  draw, summarised within each cause x age stratum. The coefficient table
  needs `cause`, `age`, `alpha`, `beta`, `gamma` and `tmrel`; several rows may
  share a cause and age group.
* `"GEMM"`: one parameter set per stratum, with the interval taken from the
  standard error of `theta`. The table needs `cause`, `age`, `theta`,
  `SE.theta`, `alpha`, `mu` and `gama`.

```{r build-cr-table}
ier <- build_cr_table(coefs, model = "IER", conc = seq(0, 20, 0.1))
names(ier)
head(ier$MEAN)
```

Alternative spellings in the coefficient file are accepted (`Causes` for
`cause`, `Age` for `age`, `zcf` for `tmrel`, `delta` for `gamma`, ...), and
`ALRI`/`All Age` are normalised to `LRI`/`ALL`. `level` sets the confidence
level of the `LOW`/`UP` sheets, and `path` writes the three sheets to an xlsx
file for use as a lookup.

## The exported functions at a glance

| Function | Purpose |
|---|---|
| `mortality()` | the analysis: attributable deaths per scenario, per cell or per domain |
| `decompose()` | two-group decomposition into `PG`/`PA`/`EXP`/`ORF` over 24 orderings |
| `aggregate_mortality()` | post hoc aggregation of a grid-level result by domain and margin |
| `aggregate_ci()` | the same for a result carrying `_MEAN`/`_UP`/`_LOW` columns |
| `write_mortality_xlsx()` | write results to a caller-supplied xlsx path |
| `build_grid_info()` | the analysis grid as a reusable, writable table |
| `cr_config()` | read a C-R model configuration |
| `cr_models()` | the accepted model names, aliases included |
| `rr_std()` | one model's lookup as a long `conc`/`endpoint`/`age`/`RR` table |
| `build_cr_table()` | a lookup built from published coefficients |
| `matchable()` | render a numeric key as a fixed-precision string |

## Where to go next

* `?mortality` -- the full argument list, including `chunk_ages`,
  `target_res` and the uncertainty arguments.
* `?decompose` -- the decomposition contract and its 24 orderings.
* `?build_grid_info` -- one grid reused across scenarios, and the
  grid-consistency check.
* `?cr_config`, `?rr_std`, `?build_cr_table` -- the concentration-response
  layer.
* `?aggregate_mortality`, `?aggregate_ci`, `?write_mortality_xlsx` -- working
  with results after the run.
* `citation("AttrMort")` and the package `README` for methodology references.
