---
title: "Getting started with AttrMort"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with AttrMort}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE
)
```

`AttrMort` estimates deaths attributable to ambient air pollution. It is
**pollutant-agnostic** (the built-in models cover PM<sub>2.5</sub>, O<sub>3</sub>
and NO<sub>2</sub>; anything else can be supplied as a lookup table or a
configuration) and **input-agnostic** (data frames, CSV/Excel files and
gridded GeoTIFF/netCDF inputs).

```{r library}
library(AttrMort)
```

This vignette walks through:

1. the data contract,
2. a tabular quick start on the example data shipped with the package,
3. domain aggregation with an uncertainty range,
4. raster inputs,
5. the concentration-response configuration and custom curves.

## The data contract

| Column | Meaning |
|---|---|
| `conc` | concentration, keyed as a character string at `dgt_conc` decimals |
| `pop` | population count of the cell |
| `age` | age stratum as character five-year groups (`"25"`, `"30"`, ...) |
| `prop` | share of the domain population in that age group |
| `endpoint` | cause label, matched against the concentration-response table |
| `mortrate` | baseline mortality, deaths per 100,000 |
| `location` | domain label used to join the domain-level tables |

Grids are joined by coordinate keys (`x`/`y`, `lon`/`lat`) rendered as
strings at `dgt_coord` decimals; wide tables carry one column per scenario and
are narrowed with `scenario =`.

## Tabular quick start

All files in `inst/extdata/` are synthetic and ship with the package:

```{r quick-start}
extdata <- system.file("extdata", package = "AttrMort")
sheet <- function(f) readxl::read_excel(file.path(extdata, f))

deaths <- mortality(
  crf       = "GEMM",
  calc_fild = sheet("grid_info.xlsx"),
  conc_real = sheet("grid_exposure.xlsx"),
  pop_total = sheet("grid_pop.xlsx"),
  age_struc = sheet("national_age_structure.xlsx"),
  mort_rate = sheet("national_mortality.xlsx"),
  scenario  = "base2015",
  mort_lvl  = "location"
)

head(deaths[, c("x", "y", "location", "ncd+lri_25")])
```

The result is one row per grid cell and one `{endpoint}_{age}` column per
stratum. `mortality()` prints an `Analysis grain:` line saying how many cells
and domains the run resolved and which PWRR branch it took.

## Domain aggregation and uncertainty

Pass `aggregate =` to sum the grid-level burdens per domain, and
`uncertain = TRUE` to attach the range built from the low/high
concentration-response tables (`conc_uncert` adds an exposure perturbation
chain on top):

```{r aggregate}
by_location <- mortality(
  crf          = "GEMM",
  calc_fild    = sheet("grid_info.xlsx"),
  conc_real    = sheet("grid_exposure.xlsx"),
  pop_total    = sheet("grid_pop.xlsx"),
  age_struc    = sheet("national_age_structure.xlsx"),
  mort_rate    = sheet("national_mortality.xlsx"),
  scenario     = "base2015",
  mort_lvl     = "location",
  aggregate    = "location",
  uncertain    = TRUE,
  conc_uncert  = 12
)

by_location[, c("location", "total", "CI_LOW", "CI_UP", "conc_pwe")]
```

`aggregate_by` keeps the endpoint and/or age breakdown
(`"total"`, `"endpoint"`, `"age"`, `"all"`). `aggregate_mortality()` and
`aggregate_ci()` do the same post hoc on a result you already have.

## Raster inputs

Rasters are aligned automatically: `conc_real` defines the analysis grid,
`pop_total` is aggregated onto it with the total preserved, and `calc_fild`
may be omitted entirely. This chunk turns one fictional country of the
example grid into two GeoTIFFs, writes them to a temporary directory, draws a
matching boundary polygon for it and runs the raster-only path:

```{r rasters}
grid <- sheet("grid_info.xlsx")
aland <- grid[grid$location == "Aland", ]
conc  <- merge(aland, sheet("grid_exposure.xlsx"), by = c("x", "y"))
pop   <- merge(aland, sheet("grid_pop.xlsx"), by = c("x", "y"))

# one 0.25 degree template; cell centres of the example grid fall on it
template <- terra::rast(xmin = 5, xmax = 30, ymin = 35, ymax = 50,
                        resolution = 0.25)
centres <- terra::xyFromCell(template, seq_len(terra::ncell(template)))
keys <- paste(round(centres[, 1], 2), round(centres[, 2], 2))
to_raster <- function(df, value) {
  r <- template
  terra::values(r) <- df[[value]][match(keys, paste(round(df$x, 2), round(df$y, 2)))]
  r
}

conc_tif <- tempfile(fileext = ".tif")
pop_tif  <- tempfile(fileext = ".tif")
terra::writeRaster(to_raster(conc, "base2015"), conc_tif)
terra::writeRaster(to_raster(pop, "base2015"), pop_tif)

boundary <- sf::st_sf(
  location = "Aland",
  geometry = sf::st_sfc(
    sf::st_polygon(list(rbind(c(5, 35), c(13, 35), c(13, 50), c(5, 50), c(5, 35)))),
    crs = 4326
  )
)

raster_deaths <- mortality(
  crf       = "GEMM",
  conc_real = conc_tif,
  pop_total = pop_tif,
  age_struc = sheet("national_age_structure.xlsx"),
  mort_rate = sheet("national_mortality.xlsx"),
  scenario  = "base2015",
  admin     = boundary,
  admin_col = "location",
  mort_lvl  = "location"
)

nrow(raster_deaths)
sum(unlist(raster_deaths[vapply(raster_deaths, is.numeric, logical(1))]))
```

`mortality()` reports when a supplied `calc_fild` does not match the raster
grid, and a multi-band raster whose scenarios do not share one validity mask
is flagged instead of silently narrowing the grid.

## Concentration-response configuration

Which lookup a model uses, which concentration column it carries and which
endpoints apply to which ages all come from a JSON configuration. The shipped
default covers the 13 built-in model names:

```{r crf}
cr_models()

cfg <- cr_config()
names(cfg$models)
cfg$models$NO2$endpoints[[1]]
```

Pass `cr_config = "<path>"` to use your own configuration, and `crf =` a
model name from it. A lookup can be a shipped `rda` table, an xlsx file or a
directory of `MEAN.csv`/`LOW.csv`/`UP.csv`; relative paths resolve against the
configuration file, not the working directory.

For a one-off pollutant/endpoint set, `build_cr_table()` turns published
coefficients into the `MEAN`/`LOW`/`UP` structure the package consumes:

```{r build-cr}
coefs <- data.frame(
  cause = "IHD", age = "25",
  alpha = 5, beta = 0.01, gamma = 0.5, tmrel = 3
)
ier <- build_cr_table(coefs, model = "IER", conc = seq(0, 20, 0.1))
names(ier)
head(ier$MEAN)
```

To hand a curve directly to `mortality(crf =)`, reshape it to the long
contract with columns `conc`, `endpoint`, `age`, `RR`; for anything reused
across analyses, register it in a configuration instead.

## Where to go next

* `?mortality` — the full argument list (`aggregate`, `uncertain`,
  `conc_uncert`, `chunk_ages`, `cr_config`).
* `?cr_config`, `?rr_std`, `?build_cr_table` — the concentration-response
  layer.
* `?build_grid_info` — reusing one grid across scenarios
  and reducing a gridded analysis to one row per domain.
* `?decompose` — driving-factor decomposition over 24 driver orderings.
* `citation("AttrMort")` and the package `README` for methodology references.
